Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-25 paper.tex

Inferring evolutionary trees while accounting for both mutation and recombination is a long-standing computational challenge. This challenge is particularly acute for SARS-CoV-2, given the large amount of data that has been shared, with over 15 million sequences in the GISAID database \citep{Shu2017-hp}. It is well known that recombination distorts phylogenies \citep{Schierup2000-fg} and affects the results of downstream analysis, e.g., inference of selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do not account for recombination \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Alternatively, genetic ancestries of recombining lineages can be represented by ancestral recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-if}, which capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.

@@

\subsection{Overview}
Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
in real time, in which recombination occurs at a low but significant rate.
The essential idea is to incrementally update the ARG each day
with the sequences collected on that day. These updates are performed by
first finding likely ``copying paths'' for each sequence in the daily
batch to the current ARG, using the ARG-based implementation of
the Li and Stephens model~\citep{Li2003-ib} introduced as
part of \texttt{tsinfer}~\citep{Kelleher2019-ba}. These copying paths
will mostly consist of a new sequence copying from a node in the ARG
with a small number of mutations, and many sequences
from a daily batch often copy from the same ARG node. The second
step is then to ``resolve'' these implied polytomies by using
standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
Such local (greedy) ARG updates inevitably lead to unparsimonious
topologies, and the third step is then to attempt to increase the
overall parsimony of the inference by making some simple topological
updates (Figure~\ref{fig:overview_sc2ts}C and D).

@@

[Quick overview of LS and ARGs/tskit computation here]
As used here, the Li and Stephens model essentially searches among
all sequences in the ARG (both sampled and inferred ancestral nodes)

The details of each of these steps are discussed in subsequent
sections.

@@

\subsection{Ancestral Recombination Graphs}
[Quick overview of ARGs and Tree Sequences/tskit. Refer to the ARG paper for
more discussion and details.]

@@

The Li and Stephens (LS) model is
a convenient probabilistic approximation to the coalescent with recombination
\citep{Hudson1983-if}, itself a generative model of a genealogical process with
recombination \citep{Li2003-ib}. Importantly, it can be implemented as an HMM
in which a focal genome is a sequence of nucleotides (observed states) that are
probabilistically emitted as an imperfect mosaic of a set of reference genomes
(hidden states). As such, an efficient HMM machinery can be employed to make
statements about parameters of interest and highlight potential sources of
error in genotyping and sequencing data. The LS model is used in a wide variety
of applications in genomics. For example, it is the engine that drives
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and
it can be used to estimate mutation rates (e.g., REF), recombination rates
\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral
sequence data \citep[e.g.][]{Palmer2019-wa}.

@@

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/imputation.pdf}
\caption{\label{fig:imputation}A schematic of the two-step procedure to impute Pango lineage for inserted, non-sample nodes. Here, three nodes (question marks) have unknown Pango lineage (A). Because the parent of node 1 is available (i.e., L2) and there is no mutation occurring on the edge connecting the two nodes, the Pango lineage of the parent is directly propagated to node 1 (B). However, if there is a mutation (red X’s), then the Pango lineage is inferred by comparison to the set of lineage-defining mutations from COVID-CG\citep{Chen2021-zc}, as in case (C). Finally, the inferred status is propagated to any non-mutant child nodes, as in case (D).
}
\end{figure}

>>>>

Inferring evolutionary trees while accounting for both mutation and
recombination is a long-standing computational challenge. This challenge is
particularly acute for SARS-CoV-2, given the large amount of data that has been
shared, with over 15 million sequences in the GISAID database
\citep{Shu2017-hp}. It is well known that recombination distorts phylogenies
\citep{Schierup2000-fg} and affects the results of downstream analysis, e.g.,
inference of selection~\citep{Anisimova2003-vr}. However, standard phylogenetic
methods do not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Alternatively,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1998-ancestral,Wong2023-efficient},
which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.

@@

\subsection{Overview}
Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
in real time, in which recombination occurs at a low but significant rate.
The essential idea is to incrementally update the ARG each day
with the sequences collected on that day. These updates are performed by
first finding likely ``copying paths'' for each sequence in the daily
batch to the current ARG, using the ARG-based implementation of
the Li and Stephens model~\citep{Li2003-ib} introduced
in \texttt{tsinfer}~\citep{Kelleher2019-ba}.
These copying paths
will mostly consist of a new sample sequence copying from a node in the ARG
with a small number of mutations, and often many samples
from a daily batch will copy from the same ARG node. The second
step is then to ``resolve'' these implied polytomies by using
standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
This greedy update strategy inevitably lead to unparsimonious
topologies, and the third step is to then increase the
overall parsimony of the inferred ARG by making some simple topological
updates (Figure~\ref{fig:overview_sc2ts}C and D).
The operations are carried out efficiently using the well-established
\texttt{tskit} library~\citep{Kelleher2018-xc,Ralph2020-efficiently},
and the resulting structure can be conveniently and interactively
analysed using the many features provided by \texttt{tskit}.

@@

@@

\subsection{Ancestral Recombination Graphs}
The term ``Ancestral Recombination Graph'' was introduced by
Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
and originally defined an alternative formulation of the coalescent
with recombination stochastic process~\citep{Hudson1983-properties}.
Subsequently, the term ARG came to be used in a more general way to
describe not just realisations of this model, but to any
recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
While there is some subtlety in the details~\citep{Wong2023-efficient},
we can think of an ARG as being any graph that encodes the
reticulate genetic ancestry of a sample of sequences under
the influence of recombination.

The ``succinct tree sequence'' is an ARG data structure
that is both general (in terms of the types of ancestry that can
be described) and computationally efficient~\citep{Wong2023-efficient}.
Originally developed to facilitate large-scale coalescent
simulations~\citep{Kelleher2016-wk}, the methods have been
extended and applied to forwards-time
simulations~\citep{Kelleher2018-xc,Haller2018-tree},
calculation of population genetics statistics~\citep{Ralph2020-efficiently}
and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
The succinct tree sequence is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations. A \emph{node}
represents a particular genome, which may be an observed sample
or an inferred genetic ancestor. The genetic inheritance between
a pair of nodes along a segment of genome is defined by
the \emph{edge} $(\ell, r, p, c)$, which states that
child node $c$ inherited its genome from parent node $p$
from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
defines the genomic position at which some variation occurs,
and a \emph{mutation} defines the site, node and derived state
that generates variation.
Given these tables, we can then efficiently construct the local
genealogical trees along the genome (arising from recombination),
and also perform a range of calculations efficiently by
reasoning about the differences between these local
trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
algorithms have led to peformance increases of several orders
of magnitude over the state-of-the-art in a range of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
Ralph2020-efficiently,Baumdicker2022-ep}.
The succinct tree sequence encoding
is also very concise, allowing, for example, for millions of
complete human genomes to be stored in a few gigabytes of
space~\citep{Kelleher2019-ba}.

The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with ARGs. Based on core functionality written
in C, it provides interfaces in C, Python and Rust. The Python interface
is based on NumPy~\citep{Harris2020-array}, and provides a convenient
platform for interactive analysis of large-scale data using, for
example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
advantage of the analysis tools in the burgeoning PyData ecosystem.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
% TODO Add more - geonomics?
applications~\citep[e.g.][]{Speidel2019-yh,Haller2019-slim}

@@

The Li and Stephens (LS) model is
a convenient probabilistic approximation to the coalescent with recombination
\citep{Hudson1983-properties}, itself a generative model of a genealogical process with
recombination \citep{Li2003-ib}. Importantly, it can be implemented as an HMM
in which a focal genome is a sequence of nucleotides (observed states) that are
probabilistically emitted as an imperfect mosaic of a set of reference genomes
(hidden states). As such, an efficient HMM machinery can be employed to make
statements about parameters of interest and highlight potential sources of
error in genotyping and sequencing data. The LS model is used in a wide variety
of applications in genomics. For example, it is the engine that drives
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and
it can be used to estimate mutation rates (e.g., REF), recombination rates
\citep[e.g.][]{Hinch2011-tz}, and selection within and across hosts in viral
sequence data \citep[e.g.][]{Palmer2019-wa}.

@@

\begin{figure} \centering
\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
\caption{\label{fig:imputation}A schematic of the two-step procedure to impute
Pango lineage for inserted, non-sample nodes. Here, three nodes (question
marks) have unknown Pango lineage (A). Because the parent of node 1 is
available (i.e., L2) and there is no mutation occurring on the edge connecting
the two nodes, the Pango lineage of the parent is directly propagated to node 1
(B). However, if there is a mutation (red X’s), then the Pango lineage is
inferred by comparison to the set of lineage-defining mutations from
COVID-CG\citep{Chen2021-zc}, as in case (C). Finally, the inferred status is
propagated to any non-mutant child nodes, as in case (D).
}
\end{figure}
